---
output:
  bookdown::word_document2:
    reference_docx: Template.docx
    fig_caption: yes
    toc: no
    number_sections: no
    df_print: kable
editor_options: 
  chunk_output_type: inline
bibliography: references.bib
link-citations: true
urlcolor: blue
linkcolor: blue
citationcolor: blue
csl: "`r here:::here('ecology.csl')`"
---

```{r setup, include=FALSE}

### This chunk of code sets up R to 

## Set options for downloading 
options(repos = c(CRAN = "http://cran.rstudio.com"))
options(timeout=60*30) #timeout downloads that last longer than 30 minutes

knitr::opts_chunk$set(
  echo = FALSE,
	message = FALSE,
	warning = FALSE,
	progress = FALSE,
	cache = FALSE,
	dpi = 300,
  fig.align = 'center'
)

# Load or install and load libraries
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  knitr, # markdown documents
  flextable, # plot tables
  bookdown, # figure numbering in markdown
  here, # easy file structure
  tidyverse, # data manipulation
  lubridate, # time
  readxl, # read excel files
  ggplot2, # figures
  patchwork, # multiple figures
  sf, # spatial data
  terra, # raster data (new pkg)
  prism, # prism data
  parallel # parallel processing
) 
terraOptions(progress=0) # suppress all progress bars in terra
cores <- parallel::detectCores() # Set number of cores for parallel processing

# Set custom plotting theme
theme_new <- function(base_size = 9,base_family = "Helvetica"){
  theme_classic(base_size = base_size, base_family = base_family) %+replace%
    theme(
      axis.line.x = element_line(color="black", linewidth = 0.25),
      axis.line.y = element_line(color="black", linewidth = 0.25),
      axis.title = element_text(size = 9),
      axis.text = element_text(colour="black", size=8),
      legend.key=element_rect(colour=NA, fill =NA),
      panel.grid = element_blank(),   
      plot.background = element_rect(fill = NA, colour = NA),
      panel.border = element_rect(fill = NA, colour = NA),
      panel.background = element_rect(fill = "white", colour = "black"), 
      strip.background = element_rect(fill = "white"),
      strip.text = element_text(size = 9)
      
    )
}
theme_set(theme_new())

set_flextable_defaults(
  font.family="Times", 
  font.size=12,
  line_spacing=1,
  padding.bottom=1,
  padding.top=1,
  text.align='center')
```

# Overview

# Methods

## Study area

## Data

### Phenology information

#### Leaf color change

1.  Early color change
2.  50% Color change
3.  All leaves changed color

#### Leaf dropping

1.  Early dropping
2.  50% leaf drop
3.  All leaves dropped

### Tree size

### Climate data

### PRISM

To characterize the climate conditions for each tree, we also obtained monthly precipitation totals, minimum temperatures, and maximum temperature 30-year normals from PRISM [@prismclimategroup2021Monthly30yearClimate] using the *prism* package [@prism].

```{r DL_PRISM}
### This chunk of code downloads gridded PRISM data for CONUS. ###
dir.create(here("Data", "Spatial"))
dir.create(here("Data", "Spatial", "PRISM"), showWarnings = FALSE) # Create directory to hold data
prism_set_dl_dir(here("Data", "Spatial", "PRISM")) # tell program to download PRISM data to the directory we just created
  
# download climate normals
get_prism_normals("ppt", "800m", annual = TRUE, keepZip = FALSE) # annual precipitation
get_prism_normals("tmean", "800m", annual = TRUE, keepZip = FALSE) # annual average temperature
get_prism_normals("tmin", "800m", mon=1, keepZip = FALSE) # minimum temperature in January
get_prism_normals("tmax", "800m", mon=7, keepZip = FALSE) # maximum temperature in Jul
```

## Data preparation

```{r importCleanData}
### FIELD DATA ###
#---- Read in excel file that links location_id field from Budburst with the Locations assigned in the field ----#
field.data <- read_xlsx(here("Data", "FieldData.xlsx"), sheet=1)
```

```{r}
### BUDBURST DATA ###
# read in budburst observation csv file
budburst <- read.csv(here("Data", "budburst_observations_1739214151.csv")) %>% 
  # convert observation_date to date object 
  mutate(date=as.Date(observation_date, format="%m/%d/%Y")) %>% 
  # pull year and day of year (doy) into separate columns
  mutate(year =lubridate::year(date), doy=lubridate::yday(date)) 
```

```{r}
#---- for each tree determine the first day---#
budburst.sub.firstdate <- budburst %>% 
  group_by(location_id) %>% 
  # for each location id calculate the minimum day
  summarize(first.doy=min(doy))

#---- for each tree determine the last day of observation---#
budburst.sub.lastdate <- budburst %>% 
  group_by(location_id) %>% 
  # for each location id calculate the minimum day
  summarize(last.doy=max(doy))

#---- for each tree determine the last date when None is reported---#
budburst.sub.dropnone <- budburst %>% 
  # limit dataset to just fall phenophase data describing leaf color
  filter(phenophase_plant_structure %in% c("Autumn Leaves Dropping")) %>% 
  # limit dataset to just observations of 50% Leaf Drop
  filter(phenophase_title == "None") %>% 
  # group by location id  
  group_by(location_id) %>% 
  # for each location id calculate the minimum day
  summarize(drop.none=max(doy))

#---- for each tree determine the first date when "Early dropping" is reported ---#
budburst.sub.drope <- budburst %>% 
  # limit dataset to just fall phenophase data describing leaf color
  filter(phenophase_plant_structure %in% c("Autumn Leaves Dropping")) %>% 
  # limit dataset to just observations of 50% Leaf Drop
  filter(phenophase_title == "Early dropping") %>% 
  # group by location id  
  group_by(location_id) %>% 
  # for each location id calculate the minimum day
  summarize(drop.early=min(doy))

#---- for each tree determine the first date when 50% of leaves had fallen ---#
budburst.sub.drop50 <- budburst %>% 
  # limit dataset to just fall phenophase data describing leaf color
  filter(phenophase_plant_structure %in% c("Autumn Leaves Dropping")) %>% 
  # limit dataset to just observations of 50% Leaf Drop
  filter(phenophase_title == "50% Leaf Drop") %>% 
  # group by location id  
  group_by(location_id) %>% 
  # for each location id calculate the minimum day
  summarize(drop.50=min(doy))

#---- for each tree determine the first date when all leaves day had dropped ---#
budburst.sub.dropall <- budburst %>% 
  # limit dataset to just fall phenophase data describing leaf color
  filter(phenophase_plant_structure %in% c("Autumn Leaves Dropping")) %>% 
  # limit dataset to just observations of All Leaf Drop
  filter(phenophase_title == "All leaves dropped") %>% 
  # group by location id  
  group_by(location_id) %>% 
  # for each location id calculate the minimum day
  summarize(drop.all=min(doy))
  
#---- combine data 50% and all leaves dropped data  ---#
dat.drop <- full_join(budburst.sub.firstdate, budburst.sub.lastdate, by="location_id")
dat.drop <- full_join(dat.drop, budburst.sub.dropnone, by="location_id")
dat.drop <- full_join(dat.drop, budburst.sub.drope, by="location_id")
dat.drop <- full_join(dat.drop, budburst.sub.drop50, by="location_id")
dat.drop <- full_join(dat.drop, budburst.sub.dropall,  by="location_id")

#---- combine Autumn Leaves Dropping data with DBH  ---#
dat.drop.field <- left_join(field.data, dat.drop, by="location_id") %>% group_by(City, Location, Tree) %>% summarise(Latitude= mean(Latitude, na.rm=T), Longitude=mean(Longitude, na.rm=T), first.doy=min(first.doy, na.rm=T), last.doy=max(last.doy, na.rm=T), drop.none=max(drop.none, na.rm=T), drop.early=min(drop.early, na.rm=T), drop.50=min(drop.50, na.rm=T), drop.all=min(drop.all, na.rm=T))

field.dat.sub <- field.data %>% select(City:Location) %>% select(-Name, -ID) %>% unique()
dat.drop.field <- left_join(field.dat.sub, dat.drop.field, by=c("City", "Location", "Tree"))

```

```{r combineWithClimate}

######### This code assigns the climate data to each unique location ###########

### Read in data from budburst and create spatial object with latitude and longitude for each location id
budburst.sf <- read.csv(here("Data", "budburst_observations_1739214151.csv")) %>% 
  # grab only location id, longitude, and latitude columns
  select(location_id, site_species_id, administrative_area_level_1, longitude, latitude) %>% 
  # remove duplicates
  unique() %>%
  # create spatial object
  st_as_sf(coords = c("longitude", "latitude"), crs = "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs", remove=F)

### Read in prism data
prism_set_dl_dir(here("Data", "Spatial", "PRISM")) # tell computer where PRISM data is stored

#### ANNUAL PPT ####
ppt.normal <- prism_archive_subset("ppt", "annual normals", resolution="800m") %>% # set what PRISM data we want to look at
  pd_to_file() %>% # get path
  rast() # read in data 
names(ppt.normal) <- "PPT"

#### ANNUAL TMEAN ####
tmean.normal <- prism_archive_subset("tmean", "annual normals", resolution="800m") %>% # set what PRISM data we want to look at
  pd_to_file() %>% # get path
  rast() # read in data 
names(tmean.normal) <- "TMEAN"

#### JANUARY TMIN ####
tmin.normal <- prism_archive_subset("tmin", "monthly normals", mon=1, resolution="800m") %>% # set what PRISM data we want to look at
  pd_to_file() %>% # get path
  rast() # read in data 
names(tmin.normal) <- "TMIN01"

#### JULY TMAX ####
tmax.normal <- prism_archive_subset("tmax",  "monthly normals", mon=7, resolution="800m") %>% # set what PRISM data we want to look at
  pd_to_file() %>% # get path
  rast() # read in data 
names(tmax.normal) <- "TMAX07"

### Overlay budburst observations and extract climate normals
budburst.sf <- extract(ppt.normal,budburst.sf, bind=T) # PPT
budburst.sf <- extract(tmean.normal,budburst.sf, bind=T) # TMEAN
budburst.sf <- extract(tmin.normal,budburst.sf, bind=T) # TMIN
budburst.sf <- extract(tmax.normal,budburst.sf, bind=T) # TMAX
budburst.climate <- budburst.sf %>% as.data.frame()

dat.drop.climate <- left_join(dat.drop, budburst.climate, by="location_id") 
write.csv(dat.drop.climate, here("Data", "Processed", "autumnleafdropXclimate.csv"), row.names=F)

### 
dat.drop.field <- dat.drop.field %>% st_as_sf(coords = c("Longitude", "Latitude"), crs = "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs", remove=F)

dat.drop.field.climate <- extract(ppt.normal,dat.drop.field, bind=T) # PPT
dat.drop.field.climate <- extract(tmean.normal,dat.drop.field.climate, bind=T) # TMIN
dat.drop.field.climate <- extract(tmin.normal,dat.drop.field.climate, bind=T) # TMIN
dat.drop.field.climate <- extract(tmax.normal,dat.drop.field.climate, bind=T) # TMAX
dat.drop.field.climate <- dat.drop.field.climate %>% as.data.frame()

write.csv(dat.drop.field.climate, here("Data", "Processed", "autumnleafdropXfieldXclimate.csv"), row.names=F)

```

## Analyses

### Were small trees more likely to drop their leaves earlier?

```{r}
# Create scatter plot of dropall by DBH
p1 <- ggplot(dat.drop.field , aes(x=DBH, y=drop.50))+geom_point()+geom_smooth(method='lm')+xlab("DBH (cm)")+ylab("First day of year\nwhere >50% of leaves dropped")
p2 <-ggplot(dat.drop.field , aes(x=DBH, y=drop.all))+geom_point()+geom_smooth(method='lm')+xlab("DBH (cm)")+ylab("First day of year\nwhere all leaves dropped")
p1 +p2+plot_annotation(tag_levels="A")+ plot_layout(guides = "collect") & theme(legend.position = 'bottom')

```

### Did trees in Fort Collins drop their leaves earlier?

```{r Q2, fig.height=3}
# Create boxplot of dropall by city
p1 <- ggplot(dat.drop.field, aes(x=City, y=drop.50))+geom_boxplot()+xlab("City")+ylab("First day of year\nwhere >50% of leaves dropped")
p2 <-ggplot(dat.drop.field, aes(x=City, y=drop.all))+geom_boxplot()+xlab("City")+ylab("First day of year\nwhere all leaves dropped")
p1 +p2+plot_annotation(tag_levels="A")
```

### Did trees from cooler climates drop their leaves earlier?

```{r Q3, fig.height=3}
dat.drop.climate <-dat.drop.climate %>% 
  # pull out just data from Colorado
  #filter(administrative_area_level_1=="CO") %>% 
  # pull only dates in Sep to Dec
  filter(drop.early > 243 & drop.50>243 & drop.all>243)



p1 <-dat.drop.climate %>% 
  select(TMIN01, drop.50) %>% #
  na.omit %>% # remove na values
  # plot
  ggplot( aes(x=TMIN01, y=drop.50))+geom_point()+geom_smooth(method="lm")+xlab("Mean minimum temperature\nin January")+ylab("First day of year\nwhere >50% of leaves dropped")
p2 <- dat.drop.climate %>% 
  select(TMIN01, drop.all) %>% #
  na.omit %>% # remove na values
  # plot
  ggplot( aes(x=TMIN01, y=drop.all))+geom_point()+geom_smooth(method="lm")+xlab("Mean minimum temperature\n in January")+ylab("Day of year where\nall leaves dropped")

p1 +p2+plot_annotation(tag_levels="A")
```

# References

::: {#refs}
:::
